A Tale of Two Dictionary Learning Problems

### 2020-08-05 manuscripts/thesis/introduction.tex
new intro

Sparse coding is a modern signal processing technique that views each of $N$ observed $n$-dimensional signal samples as a (noisy) linear combination of at most $k$ elementary waveforms drawn from some unknown ``dictionary" of size $m \ll N$ (see \cite{Zhang15} for a comprehensive review). 
%\IEEEPARstart{A}{}common modern approach to pattern analysis in signal processing is to view each of $N$ observed $n$-dimensional signal samples as a (noisy) linear combination of at most $k$ elementary waveforms drawn from some unknown ``dictionary" of size $m \ll N$ (see \cite{Zhang15} for a comprehensive review). 
Optimizing dictionaries subject to this and related sparsity constraints has revealed seemingly characteristic sparse structure in several signal classes of current interest (e.g., in vision \cite{wang2015sparse}). 
Of particular note are the seminal works in the field \cite{Olshausen96, hurri1996image, bell1997independent, van1998independent}, which discovered that dictionaries optimized for coding small patches of ``natural" images share qualitative similarities with linear filters estimated from response properties of simple-cell neurons in mammalian visual cortex. Curiously, these waveforms (e.g., ``Gabor'' wavelets) appear in dictionaries learned by a variety of algorithms trained over different natural image datasets, suggesting that learned features in natural signals may, in some sense, be canonical \cite{donoho2001can}.

Motivated by these discoveries and more recent work relating compressed sensing \cite{eldar2012compressed} to a theory of information transmission through random wiring bottlenecks in the brain \cite{Isely10}, this thesis addresses when dictionaries for sparse representation are indeed identifiable from data. Answers to this question may also have implications in practice wherever an appeal is made to latent sparse structure of data (e.g., forgery detection \cite{hughes2010, olshausen2010applied}; brain recordings \cite{jung2001imaging, agarwal2014spatially, lee2016sparse}; and gene expression \cite{wu2016stability}). 
While several algorithms have recently been proposed to provably recover unique dictionaries under specific conditions (see \cite[Sec.~I-E]{Sun16} for a summary of the state-of-the-art), few theorems can be invoked to justify the consistency of inference under this model of data more broadly. To my knowledge, a universal guarantee of the uniqueness and stability of learned dictionaries and the sparse representations they induce over noisy data has yet to appear in the literature.

>>>>

%Sparse representation modeling is an approach to signal processing that seeks to describe signals as linear combinations of a few elementary waveforms selected from a pre-specified ``dictionary".

It is a longstanding practice in the field of signal processing to describe signals as linear combinations of elementary waveforms from a pre-specified ``dictionary". When this dictionary forms a basis for the signal space, every signal has a unique decomposition into these atomic components. In the simplest case, the basis is orthonormal and the representational coefficient scaling a given elementary waveform is merely the inner product of that waveform with the signal. %When non-orthogonal, the coefficients are computed by taking the inner product with respect to the components of the dictionary inverse.

Until recently, bases have been the default form of signal representation due largely to their simplicity. For many signal analysis tasks, however, there is no one basis expressive enough to reveal clearly all of the relevant features of the signal. For example, a signal can be decomposed into its constituent frequencies via the Fourier transform, a linear change of basis. If our signal can be either a sine wave or a delta function, neither the standard basis nor the Fourier basis can capture one case as effectively as it can the other.

The need for greater freedom of expression led to the development of redundant signal representations utilizing overcomplete dictionaries containing more atoms than there are dimensions of the signal. In this case, there are infinitely many ways in which a signal may be decomposed into its constituent components, and the intention to seek the most informative such representation as measured by some task-specific cost function.

A popular approach to the design of overcomplete dictionaries has been to seek one with respect to which every signal in the signal class of interest has a sparse representation; that is, it can be represented, or at least well-approximated, as a combination of only a few dictionary elements from the bunch. Concluding our example, the union of the standard basis with the Fourier basis is an over-complete dictionary with respect to which both sines and delta functions (or any superposition thereof) have the sparsest possible representations as combinations of its elementary waveforms.

Early approaches to sparse coding assumed, as was assumed in our recurring example, a model of the signal class from which a suitable sparsifying dictionary could then be derived. While such dictionaries are typically characterized by an analytic formulation and a fast implicit implementation, these models tend to be over-simplistic when applied to natural phenomena.

In contrast, an alternative approach to dictionary design is conditioned on the assumption that the structure of complex natural phenomena can be more accurately extracted directly from a training dataset of signals, a process referred to as dictionary learning. In the seminal work \cite{Olshausen96}, the authors trained a dictionary for sparse representation of small image patches collected from a number of images of the natural environment. Remarkably, the relatively simple algorithm they applied produced dictionary elements sharing qualitative similarities with linear filters estimated from response properties of simple-cell neurons in mammalian visual cortex, which until then were more weakly described by the analytic Gabor filters (see also \cite{hurri1996image, bell1997independent, van1998independent}). Curiously, these waveforms (e.g., Gabor'-like wavelets) appear in dictionaries learned by a variety of algorithms trained over different natural image datasets, suggesting that learned features in natural signals may, in some sense, be canonical \cite{donoho2001can}.

% C&P: A key contribution to the area of dictionary learning was provided by Olshausen and Field in 1996 [34]. In their widely celebrated paper, the authors trained a dictionary for sparse representation of small image patches collected from a number of natural images. With relatively simple algorithmic machinery, the authors were able to show a remarkable resultVthe trained atoms they obtained were incredibly similar to the mammalian simple-cell receptive fields, which until then were only weakly explained via Gabor filters. The finding was highly motivating to the sparse representation community, as it demonstrated that the single assumption of sparsity could account for a fundamental biological visual behavior. Also, the results demonstrated the potential in example-based methods to uncover elementary structures in complex signal data.

Sparse coding is a modern signal processing technique that views each of $N$ observed $n$-dimensional signal samples as a (noisy) linear combination of at most $k$ elementary waveforms drawn from a ``dictionary" of size $m \ll N$ (see \cite{Zhang15} for a comprehensive review). 
%\IEEEPARstart{A}{}common modern approach to pattern analysis in signal processing is to view each of $N$ observed $n$-dimensional signal samples as a (noisy) linear combination of at most $k$ elementary waveforms drawn from some unknown ``dictionary" of size $m \ll N$ (see \cite{Zhang15} for a comprehensive review). 
Optimizing dictionaries subject to this and related sparsity constraints has revealed seemingly characteristic sparse structure in several signal classes of current interest (e.g., in vision \cite{wang2015sparse}). 
Of particular note are the seminal works in the field \cite{Olshausen96, hurri1996image, bell1997independent, van1998independent}, which discovered that dictionaries optimized for coding small patches of ``natural" images share qualitative similarities with linear filters estimated from response properties of simple-cell neurons in mammalian visual cortex. Curiously, these waveforms (e.g., ``Gabor'' wavelets) appear in dictionaries learned by a variety of algorithms trained over different natural image datasets, suggesting that learned features in natural signals may, in some sense, be canonical \cite{donoho2001can}.

Motivated by these discoveries and more recent work relating compressed sensing \cite{eldar2012compressed} to a theory of information transmission through random wiring bottlenecks in the brain \cite{Isely10}, this thesis concerns itself with the following question: when is a dictionary indeed identifiable from data? Answers to this question may also have implications in practice wherever an appeal is made to latent sparse structure of data (e.g., forgery detection \cite{hughes2010, olshausen2010applied}; brain recordings \cite{jung2001imaging, agarwal2014spatially, lee2016sparse}; and gene expression \cite{wu2016stability}). 
While several algorithms have recently been proposed to provably recover unique dictionaries under specific conditions (see \cite[Sec.~I-E]{Sun16} for a summary of the state-of-the-art), few theorems can be invoked to justify the consistency of inference under this model of data more broadly. To my knowledge, a universal guarantee of the uniqueness and stability of learned dictionaries and the sparse representations they induce over noisy data has yet to appear in the literature.
